\subsection{WSDL: an extra complication}
\label{objections:wsdl-gen}

The role of an interface definition language (IDL) has always been
twofold:

\begin{enumerate}
\item Firstly, an IDL allows the creation of a definition of \emph{the
interface} of the remote system, independent of any particular
implementation, programming language or environment. This is
``interface'' in the sense of \emph{the implementation independent
signature of the service}, and does not imply that an implementation
language needs an explicit notion of interfaces. The interface is
inherently implementation independent, and can be frozen or carefully
managed with respect to versioning.

\item Secondly, the act of writing an IDL description forces the author to
define the system in terms of the portable datatypes and operations available in
the restricted language of the IDL. This can effectively guarantee portability,
and is a significant improvement over interfaces defined in the implementation
languages themselves, which invariably contain constructs which are not
portable.

\end{enumerate}

IDLs have many advantages for creating interoperable systems, yet the
generally accepted practice for working with JAX-RPC discards all
these notions. Instead of generating implementation classes from WSDL,
the WSDL description is usually generated from the implementation
classes using tools leveraging Java's Reflection API. We shall term
this process \emph{contract-last development}.

This has the following consequences.

\begin{itemize}

\item
    
There is no way to ensure that a service's published interface remains constant
over time. Every redeployment of the service or may change the classes, and
hence the contract.

\item

Some aspects of the interface a service provides are not amenable to
extraction simply from signatures of its constituent classes and
methods. A given method might, for example, choose to extract
attachments from messages programmatically, without declaring that
attachment in the signature.

\item

No warnings about portability issues are available before deployment
occurs. When defining a service using an IDL, the author typically
knows when there are problems as the IDL will not compile. Yet with
contract-last development, everything may well seem to work until the
service goes live and a customer using a different language attempts
to import the WSDL and invoke the service.
    
\end{itemize}

The alternative to contract-last development is clearly \emph{contract-first
development}. Although this is the better approach from the perspective
of portability and interface stability, web service developers are not
pushed in this direction.

One of the underlying causes of this has to be the sheer complexity of
XML Schema and WSDL. The XSD type system bears minimal resemblance to
that of current object oriented languages, and WSDL itself is
over-verbose and under-readable. As evidence of this, consider the
broad variety of products that aim to make authoring XSD and WSDL
documents easier, and recall that such products were never necessary
in the IDL-era of distributed systems programming.

%We note in passing that that REST systems \cite{fielding:rest} tend not to
%make use of WSDL, even though it is theoretically possible. Instead
%they resort to their XML type language of choice and quality
%human-readable documentation. This would appear to be sub-optimal, yet
%REST is growing in popularity, despite (or perhaps because of) the
%lack of WSDL integration.

Returning to the desiderata for SOAP, following a contract-last process
sacrifices interoperability for ease of service development. Perhaps
WSDL is not the appropriate language for describing SOAP services (we
are certainly not enthused about it), yet the sole solution being
advocated is not a major undertaking to fix WSDL's core flaws, it is
to continue to encourage developers to hand over to their SOAP stacks
the challenge of deriving a stable and portable service interface from
the inherently unstable and unportable service implementation.

We are not proposing any changes to WSDL, merely mourning the fact
that its over-complexity discourages contract-driven
development more aggressively than any previous IDL ever did. We do
observe that once the type declarations of a service have been moved
into their own document, WSDL becomes much more manageable and this is
a pattern of service definition which we strongly encourage.
